---
title: "Reproducible Research Course Project 2"
author: "Benjamin Estrade"
date: "24 July 2018"
output: 
  html_document: 
    keep_md: yes
---

Synopsis
========














Project Goal
============

Use the data to answer some basic questions about severe weather events.

Questions:
1. Across the United States, which types of events (as indicated in the <span style="color:red" face="KaTeX_Typewriter">EVTYPE</span> variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?


Data Processing
===============

Details about project
---------------------

Copy of the data can be found in the [GitHub directory](https://github.com/benegd/reproducible-research-course-project-2).

Data retrived from the [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

This project is part of the data science specialization course by John Hopkins University hosted by [Coursera](https://www.coursera.com/).

Consult README.md for more information


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

System info
-----------

```{r}
sessionInfo()
```


Downloading Data from orginal source
------------------------------------

``` {r}
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zippath <- "./ProjectData/Storm_data.csv.bz2"
projectdatapath <- "./ProjectData"
if(!file.exists(zippath)){
        if(!dir.exists(projectdatapath)){
               dir.create(projectdatapath) 
        }
        download.file(fileurl, zippath)
}
        
```


Loading data into R
-------------------
``` {r}
stormdata <- read.csv(zippath)
head(stormdata)
summary(stormdata)
str(stormdata)
```




Results
=======












